Glyph-maps for Visually Exploring Temporal Patterns in Climate Data and Models

### 2012-03-06 revision/paper.tex
done with large data

Climate data is usually large: many locations in space, and many points in time, which can make it arduous to work with computationally. The glyph-map calculations are as efficient as possible computationally, being linear in the number of spatial locations. It is possible that the calculations could also be paralellized for massive amounts of climate model data. Visualizing large data also hits a perceptual boundary.  There is a limit to the resolution of human visual system \citep{Kr12} which will prevent perfect perception of large spatial domains or seasonality of many years of data. Exploring large data requires reducing it to a resolution that can be reasonably absorbed visually.

Figure~\ref{fig:gistemp-pred} shows the glyph-map for the GISTEMP dataset, which is more realistically sized than the EXPO data. Glyphs at each location represent temperature from 1880--2011, starting with Jan 1880 at 12 o'clock and proceeding clockwise to 2011 at 11:59. The most important feature to notice is the pattern of missing values. Most glyphs do not complete the full circle, typically missing points between 12 and 1, suggesting missing values in the earliest part of the data. Another important feature is variability, shown by the thickness of the circle. Some locations, particularly in the north western mountain region, have large variations in temperature.

@@

The raw data shows us missing values and gives some sense of variability, but gives no information about long-term trend. Statistical models are useful for pre-processing climate data, to decompose data into long-term trend, seasonal effects and error, which is vital for making the problem both computationally and visually tractable. Trend is examined by fitting a linear model to each location and displaying predicted values in Figure~\ref{fig:gistemp-pred}. This plot shows long-term (de-seasonalized) temperature trends for 1950--2010 (the limited time period used due to missing data). Most locations have increasing temperature over this time period. Only a few locations show a decline, and these few locations tend to be different from their neighbors, and on the edge of the data collection region, indicating that there might be isolated data problems. The rate of increase differs across regions: steeper inclines in the north and western mountain regions and across the Great Lakes, and flatter inclines in the southeastern USA.

>>>>

Climate data is usually large: many locations in space, and many points in time, which can make it arduous to work with computationally. The glyph-map calculations are as efficient as possible computationally, being linear in the number of spatial locations. It is possible that the calculations could also be parallelized for massive amounts of climate model data. Visualizing large data also hits a perceptual boundary.  There is a limit to the resolution of human visual system \citep{Kr12} which will prevent perfect perception of large spatial domains or seasonality of many years of data. Exploring large data requires reducing it to a resolution that can be reasonably absorbed visually.

Figure~\ref{fig:gistemp-pred} shows the glyph-map for the GISTEMP dataset, which is more realistically sized than the EXPO data. Glyphs at each location represent temperature from 1880--2011, starting with Jan 1880 at 12 o'clock and proceeding clockwise to 2011 at 11:59. The most important feature to notice is the pattern of missing values. Most glyphs do not complete the full circle, typically missing points between 12 and 1, suggesting missing values in the earliest part of the data. Another important feature is variability, shown by the thickness of the circle. Some locations, particularly in the north and western mountain region, have large variations in temperature.

@@

The raw data shows us missing values and gives some sense of variability, but gives no information about long-term trend. Statistical models are useful for pre-processing climate data, to decompose data into long-term trend, seasonal effects and error, which is vital for making the problem both computationally and visually tractable. Trend is examined by fitting a linear model to each location and displaying predicted values in Figure~\ref{fig:gistemp-pred}. This plot shows long-term (de-seasonalized) temperature trends for 1950--2010. (The reduced time period is used to eliminate the influence of the missing data values in early years.) Most locations have increasing temperature over this time period. Only a few locations show a decline, and these few locations tend to be different from their neighbors, and on the edge of the data collection region, indicating that there might be isolated data problems. The rate of increase differs across regions: steeper inclines in the north and western mountain regions and across the Great Lakes, and flatter inclines in the southeastern USA.
